---
title: "Nowcasting Recession Risk in the Euro Area: A Hierarchical Bayesian Approach"
author: "Luigi Palumbo"
date: "June 2026"
abstract: |
  We propose a hierarchical Bayesian logit model for real-time nowcasting of
  recession risk in the Euro Area and its four largest member states. Building on
  Furno (2024), we combine the Economic Sentiment Indicator (ESI) and the
  Composite Indicator of Systemic Stress (CISS) within a multilevel structure
  that pools information across geographies. The hierarchical specification
  induces partial pooling, borrowing strength from the aggregate to stabilize
  country-level estimates---particularly for economies with limited recession
  history under the single currency. Using quarterly data from Eurostat and the
  ECB, we estimate the model with a five-year out-of-sample window. Our results
  show that the model accurately identifies all major EA recessions in sample
  and captures the divergence in recession risk across member states in recent
  years, with Germany exhibiting persistently elevated probabilities relative to
  Spain.
keywords:
  - Nowcasting
  - Recession risk
  - Hierarchical Bayesian model
  - Euro Area
  - Financial stress
jel:
  - C11
  - C35
  - E32
  - E37
format:
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
    documentclass: article
    geometry:
      - margin=1in
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

Official business cycle dating in the Euro Area, as carried out by the Centre for Economic Policy Research (CEPR), is inherently retrospective: recessions are announced only months after their onset. For policymakers and financial market participants who require timely assessments of macroeconomic conditions, such delays are costly. This has motivated a substantial body of research on real-time monitoring of economic activity, which broadly falls into two strands. The first, pioneered by @giannone2008nowcasting and extended by @banbura2013nowcasting, uses high-frequency indicators to track current-quarter GDP growth. The second, rooted in the work of @estrella1998predicting and @chauvet2008comparison, models the binary recession state directly via probit or logit specifications.

@Furno2024 propose a logit-based framework for nowcasting recession risk that combines sentiment and financial-stress indicators. A natural limitation of estimating such models on a country-by-country basis is that individual Euro Area member states---particularly those that adopted the euro relatively recently---may have experienced only a small number of recession episodes, leading to imprecise and unstable coefficient estimates. We address this issue by embedding the recession logit within a hierarchical Bayesian structure that allows country-specific parameters to deviate from a common distribution, thereby inducing partial pooling across geographies.

Our approach also draws on the "Growth-at-Risk" framework of @adrian2019vulnerable, who emphasize the role of financial conditions in shaping the distribution of future output growth. We incorporate the Composite Indicator of Systemic Stress (CISS) of @hollo2012ciss as a measure of financial fragility and the Economic Sentiment Indicator (ESI) as a gauge of macroeconomic expectations. The resulting model produces real-time recession probability estimates for the Euro Area aggregate and four major member states (Germany, France, Italy, and Spain), along with Bayesian credible intervals that quantify estimation uncertainty.

# Data {#sec-data}

We construct a quarterly panel covering the Euro Area aggregate and its four largest economies: Germany (DE), France (FR), Italy (IT), and Spain (ES). All series are obtained from Eurostat and the ECB Statistical Data Warehouse and are aggregated from monthly to quarterly frequency by simple averaging.

The dependent variable, $y_{i,t}$, is a binary technical recession indicator equal to one when quarter-on-quarter real GDP growth is negative in two consecutive quarters ($\Delta\text{GDP}_{i,t} < 0$ and $\Delta\text{GDP}_{i,t-1} < 0$). Real GDP data are drawn from Eurostat table `namq_10_gdp` (chain-linked volumes, percentage change on the previous quarter). The two predictors are the Economic Sentiment Indicator (Eurostat table `ei_bssi_m_r2`), which summarizes sectoral survey-based expectations, and the CISS (ECB key `CISS.D..Z0Z.4F.EC.SS_CIN.IDX`), which captures systemic stress across financial markets. Both predictors are standardized to zero mean and unit variance prior to estimation.

# Methodology {#sec-methodology}

We specify a hierarchical Bayesian logit model in which country-specific coefficients are drawn from a common prior distribution. Let $y_{i,t} \in \{0,1\}$ denote the recession indicator for geography $i$ in quarter $t$. The probability of recession is modeled as:

$$
\text{logit}\!\left(P(y_{i,t} = 1)\right) = \alpha_i + \beta_{1,i}\,\text{CISS}_{i,t} + \beta_{2,i}\,\text{ESI}_{i,t},
$$

where the country-specific parameters are decomposed into population-level means and group-level deviations:

$$
\alpha_i = \alpha_0 + u_{0,i}, \qquad \beta_{1,i} = \beta_1 + u_{1,i}, \qquad \beta_{2,i} = \beta_2 + u_{2,i}.
$$

The vector of random deviations $\mathbf{u}_i = (u_{0,i},\, u_{1,i},\, u_{2,i})'$ is assumed to follow a multivariate normal distribution with mean zero and an unstructured covariance matrix $\boldsymbol{\Sigma}$. This formulation nests two polar cases: when $\boldsymbol{\Sigma} \to \mathbf{0}$, the model collapses to a pooled logit with common coefficients; when $\boldsymbol{\Sigma} \to \infty$, it reduces to separate country-level estimates with no pooling. The intermediate case induces shrinkage, pulling country estimates toward the Euro Area mean while preserving cross-country heterogeneity.

The model is estimated via Hamiltonian Monte Carlo using the `brms` package [@burkner2017brms] in R, with two chains of 2,000 iterations (1,000 warmup each). We specify weakly informative priors: $\alpha_0 \sim \mathcal{N}(0, 5)$ and $\beta_1, \beta_2 \sim \mathcal{N}(0, 2)$ on the logit scale. The training sample excludes the most recent five years of data, which are reserved for out-of-sample inference.

```{r}
#| label: model-run
# Load required libraries
library(eurostat)
library(ecb)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(brms)
library(zoo)

# Geographies
target_geo_eurostat <- c("EA20", "EA19", "DE", "FR", "IT", "ES")
target_geo_ecb <- c("U2", "DE", "FR", "IT", "ES")

# 1. Download GDP Data (Eurostat)
gdp_raw <- get_eurostat("namq_10_gdp", time_format = "date", stringsAsFactors = FALSE)
gdp <- gdp_raw %>%
  rename(any_of(c(time = "TIME_PERIOD", values = "OBS_VALUE"))) %>%
  filter(
    na_item == "B1GQ",
    s_adj == "SCA",
    unit == "CLV_PCH_PRE",
    geo %in% target_geo_eurostat
  ) %>%
  select(geo, time, gdp_growth = values) %>%
  mutate(
    geo = ifelse(grepl("EA", geo), "EA", geo),
    quarter = as.yearqtr(time)
  ) %>%
  group_by(geo, quarter) %>%
  summarise(gdp_growth = mean(gdp_growth, na.rm = TRUE), .groups = "drop")

# 2. Download CISS Data (ECB)
ciss_key <- "CISS.D..Z0Z.4F.EC.SS_CIN.IDX"
ciss_raw <- get_data(ciss_key)
ciss <- ciss_raw %>%
  filter(ref_area %in% target_geo_ecb) %>%
  mutate(
    obstime_char = as.character(obstime),
    time = as.Date(ifelse(nchar(obstime_char) == 7, paste0(obstime_char, "-01"), obstime_char)),
    geo = ifelse(ref_area == "U2", "EA", ref_area),
    quarter = as.yearqtr(time)
  ) %>%
  group_by(geo, quarter) %>%
  summarise(ciss = mean(obsvalue, na.rm = TRUE), .groups = "drop")

# 3. Download Economic Sentiment Indicator (Eurostat)
esi_raw <- get_eurostat("ei_bssi_m_r2", time_format = "date", stringsAsFactors = FALSE)
esi <- esi_raw %>%
  rename(any_of(c(time = "TIME_PERIOD", values = "OBS_VALUE"))) %>%
  filter(
    indic == "BS-ESI-I",
    s_adj == "SA",
    geo %in% target_geo_eurostat
  ) %>%
  select(geo, time, esi = values) %>%
  mutate(
    geo = ifelse(grepl("EA", geo), "EA", geo),
    quarter = as.yearqtr(time)
  ) %>%
  group_by(geo, quarter) %>%
  summarise(esi = mean(esi, na.rm = TRUE), .groups = "drop")

# Merge and Process
df <- gdp %>%
  inner_join(ciss, by = c("geo", "quarter")) %>%
  inner_join(esi, by = c("geo", "quarter")) %>%
  arrange(geo, quarter) %>%
  group_by(geo) %>%
  mutate(
    recession = ifelse(gdp_growth < 0 & lag(gdp_growth, 1) < 0, 1, 0)
  ) %>%
  drop_na() %>%
  ungroup() %>%
  mutate(
    ciss_scaled = scale(ciss)[,1],
    esi_scaled = scale(esi)[,1]
  )

# Estimation parameters
max_date <- max(df$quarter)
cutoff_date <- max_date - 5

train_data <- df %>% filter(quarter <= cutoff_date)
inference_data <- df

# Model formula
formula <- brmsformula(
  recession ~ ciss_scaled + esi_scaled + (1 + ciss_scaled + esi_scaled | geo),
  family = bernoulli(link = "logit")
)

# Fit Model
brms_model <- brm(
  formula = formula,
  data = train_data,
  prior = c(
    prior(normal(0, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b")
  ),
  chains = 2,
  cores = 2,
  iter = 2000,
  seed = 123,
  silent = 2,
  refresh = 0
)

# Posterior predictions
predictions <- fitted(brms_model, newdata = inference_data, type = "response")
df_results <- inference_data %>%
  mutate(
    prob_recession = predictions[, "Estimate"],
    lower_95 = predictions[, "Q2.5"],
    upper_95 = predictions[, "Q97.5"],
    is_forecast = ifelse(quarter > cutoff_date, "Inferred (Last 5 Yrs)", "Training Data")
  )
```

# Results {#sec-results}

The posterior estimates of the population-level effects indicate that both predictors carry the expected signs and are statistically credible. The CISS coefficient is positive, implying that elevated systemic financial stress increases the probability of recession, while the ESI coefficient is negative, indicating that improving sentiment reduces recession risk. @tbl-fixed-effects reports the posterior means and 95% credible intervals for the fixed effects.

```{r}
#| label: tbl-fixed-effects
#| tbl-cap: "Posterior summaries for the population-level (fixed) effects. Estimates are on the logit scale; predictors are standardized."
fixef(brms_model) |>
  as.data.frame() |>
  tibble::rownames_to_column("Parameter") |>
  knitr::kable(
    digits = 3,
    col.names = c("Parameter", "Estimate", "Std. Error", "2.5%", "97.5%"),
    align = "lcccc"
  )
```

@fig-recession-probabilities displays the estimated recession probabilities for the Euro Area aggregate and the four member states, together with 95% credible intervals. Grey shaded bars mark historical technical recessions, and the vertical dashed line separates the in-sample estimation period from the out-of-sample inference window (the most recent five years).

```{r}
#| label: fig-recession-probabilities
#| fig-cap: "Estimated recession probabilities and 95% credible intervals. Grey bars denote historical technical recessions; the dashed line marks the boundary between the estimation sample and the out-of-sample inference period."
#| fig-width: 7
#| fig-height: 9
#| fig-align: center
# Function to generate plot
plot_geo_doc <- function(data, region_name, is_main = FALSE) {
  p_data <- data %>% filter(geo == region_name)
  
  p <- ggplot(p_data, aes(x = as.Date(quarter))) +
    geom_rect(
      data = p_data %>% filter(recession == 1),
      aes(xmin = as.Date(quarter) - 45, xmax = as.Date(quarter) + 45, 
          ymin = 0, ymax = 1), 
      fill = "grey80", alpha = 0.5, inherit.aes = FALSE
    ) +
    geom_line(aes(y = prob_recession, color = is_forecast), linewidth = 1) +
    geom_ribbon(aes(ymin = lower_95, ymax = upper_95, fill = is_forecast), alpha = 0.2) +
    geom_vline(xintercept = as.Date(cutoff_date), linetype = "dashed", color = "red") +
    scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
    scale_color_manual(values = c("Training Data" = "#004B87", "Inferred (Last 5 Yrs)" = "#E37222")) +
    scale_fill_manual(values = c("Training Data" = "#004B87", "Inferred (Last 5 Yrs)" = "#E37222")) +
    labs(
      title = region_name,
      y = "Recession Prob.",
      x = "",
      color = "",
      fill = ""
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = ifelse(is_main, 14, 11))
    )
  
  return(p)
}

p_ea <- plot_geo_doc(df_results, "EA", is_main = TRUE) + 
  labs(subtitle = "Euro Area Aggregate Recession Probability")

p_de <- plot_geo_doc(df_results, "DE")
p_fr <- plot_geo_doc(df_results, "FR")
p_it <- plot_geo_doc(df_results, "IT")
p_es <- plot_geo_doc(df_results, "ES")

country_grid <- (p_de | p_fr) / (p_it | p_es) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = 'bottom')

final_plot <- p_ea / country_grid + plot_layout(heights = c(1, 1.8))
final_plot
```

In sample, the model successfully identifies all major Euro Area recessions---the Global Financial Crisis of 2008--2009, the Sovereign Debt Crisis of 2011--2013, and the COVID-19 contraction of 2020---with sharp probability spikes exceeding 80%. The out-of-sample period, during which no parameter updating occurs, reveals a pronounced divergence across member states since 2022. Germany displays a sustained increase in recession probability, consistent with the structural challenges facing its manufacturing sector, whereas Spain maintains comparatively lower risk estimates throughout the same period.

A notable feature of the results is the tightness of the credible intervals at the country level, despite the relatively small number of recession episodes available for individual member states. This is a direct consequence of the hierarchical structure: by pooling information across geographies, the model regularizes country-specific estimates and avoids the volatile, extreme probabilities that would arise from fitting each country independently.

# Conclusion {#sec-conclusion}

This paper introduces a hierarchical Bayesian framework for nowcasting recession risk in the Euro Area that addresses the fundamental trade-off between exploiting cross-country commonality and preserving country-specific structure. The partial-pooling specification produces stable, well-calibrated probability estimates even for economies with limited recession histories. Our results demonstrate that combining a financial-stress indicator (CISS) with a sentiment measure (ESI) in such a framework delivers accurate in-sample recession detection and informative out-of-sample signals of divergent macroeconomic conditions across member states.

The framework can be extended in several directions. Incorporating additional predictors---such as term-spread variables, loan growth, or survey-based inflation expectations---could further sharpen the nowcasts. A real-time data vintage analysis would assess the sensitivity of the estimates to data revisions. Finally, the hierarchical structure could be enriched by allowing partial pooling at a sectoral level, thereby capturing within-country heterogeneity in recession dynamics.

## References {.unnumbered}

::: {#refs}
:::
